On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-03-27 manuscripts/PNAS/chazthm_pnas.tex
chazthm_pnas.tex

Motivated by these discoveries and earlier work in the theory of neural communication \cite{Coulter10, Isely10}, we address when dictionaries and sparse representations are uniquely determined by data.  Answers to this question also have other real-world implications.  For example, a sparse coding analysis of local painting style can be used for forgery detection \cite{hughes2010, Olshausen10}, but only if sparse components of the artwork are independent of implementation idiosyncrasies. Fortunately, algorithms with proven recovery of generating dictionaries under certain 
conditions have recently been proposed (see \cite[Sec.~I-E]{Sun16} for a summary of the state-of-the-art).  However,
there are few general theorems explaining the general phenomenon of uniqueness.

@@

Note that any particular solution $(\mathbf{B}, \mathbf{\hat x}_1 \ldots, \mathbf{\hat x}_N)$ to this problem gives rise to an orbit of other solutions $(\mathbf{BPD}, \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\hat x}_1 \ldots, \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\hat x}_N)$, where $\mathbf{P}$ is any permutation matrix and $\mathbf{D}$ any invertible diagonal.  
We therefore consider whether solutions to Problem~\ref{InverseProblem} are unique only up to this ambiguity.  More specifically, 
we seek conditions for when sparsely coded datasets have a stable unique representation.

@@

%  We introduce the following terminology to handle $\eta > 0$. 
\begin{definition}\label{maindef}
Fix $Y = \{ \mathbf{y}_1, \ldots, \mathbf{y}_N\} \subset \mathbb{R}^n$. We say $Y$ has a \textbf{$k$-sparse representation in $m$ dimensions} if there exists a matrix $\mathbf{A}$ and $k$-sparse $\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$ such that $\mathbf{y}_i = \mathbf{A}\mathbf{x}_i$ for all $i$. %Equivalently, $\mathbf{Y} = \mathbf{A}\mathbf{X}$ for $\mathbf{Y}$ and $\mathbf{X}$ having the $\mathbf{y}_i$ and $\mathbf{x}_i$ as their columns, respectively. 
This representation is \textbf{stable} if for every $\delta_1, \delta_2 \geq 0$, there exists $\varepsilon = \varepsilon(\delta_1, \delta_2) \geq 0$ (with $\varepsilon > 0$ when  $\delta_1, \delta_2 > 0$) such that if a matrix $\mathbf{B}$ and $k$-sparse $\mathbf{\hat x}_1, \ldots, \mathbf{\hat x}_N \in \mathbb{R}^m$ satisfy:
\begin{align*}
\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \varepsilon, \ \ \text{for all $i$},
\end{align*}
%
then there is some permutation matrix $\mathbf{P}$ and invertible diagonal matrix $\mathbf{D}$ such that for all $i, j$:
\begin{align}\label{def1}
\|\mathbf{A}_j - (\mathbf{BPD})_j\|_2 \leq \delta_1 \ \ \text{and} \ \ \|\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\hat x}_i\|_1 \leq \delta_2.
\end{align}
\end{definition}
% Kato mentions a distance between n-tuples that minimizes the maximum difference over all possible orderings of one of the n-tuples.
% Make just 'sparse representation in m-dimensions? (i.e. with k unspecified?)

To see how Definition \ref{maindef} directly relates to Problem \ref{InverseProblem}, suppose that $Y$
% \mbox{$Y = \{\mathbf{A} \mathbf{x}_1, \ldots, \mathbf{A}\mathbf{x}_N\}$} 
has a stable $k$-sparse representation in $m$ dimensions and fix $\delta_1, \delta_2$ to be the desired accuracy in \eqref{def1}. Consider now any dataset $Z$ generated as in \eqref{LinearModel} with $\eta \leq \frac{1}{2} \varepsilon(\delta_1, \delta_2)$. Then, any dictionary $\mathbf{B}$ and $k$-sparse $\mathbf{\hat x}_1, \ldots, \mathbf{\hat x}_N$ solving Problem \ref{InverseProblem} are necessarily close to the original matrix $\mathbf{A}$ and codes $\mathbf{x}_i$ (i.e., satisfy \eqref{def1}).

@@

[*** TODO: Combine with Theorem 3 ***]
\begin{theorem}\label{DeterministicUniquenessTheorem}
Fix integers $1 \leq p \leq m$, $n$, and $k < m \leq m' \leq \frac{\ell m - (p-1)}{\ell - 1}$. Suppose $\mathbf{A}~\in~\mathbb{R}^{n \times m}$ satisfies \eqref{SparkCondition} and that \mbox{$\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$} include at least \mbox{$(k-1){m \choose k}+1$} vectors in general linear position supported on each set in some $\ell$-regular $E \subseteq {[m] \choose k}$ satisfying the SIP. Then there exists a constant $C_1 > 0$ for which the following holds for all $\varepsilon < L_2(\mathbf{A}) / C_1$. %\frac{L_2(\mathbf{A})}{\sqrt{2}}C_1^{-1}$:

Every matrix $\mathbf{B} \in \mathbb{R}^{n \times m}$ for which for some $k$-sparse $\mathbf{\hat x}_1, \ldots, \mathbf{\hat x}_N \in \mathbb{R}^{m}$ we have \mbox{$\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \varepsilon$} for all $i$ necessarily satisfies:
\begin{align}\label{Cstable}
\|\mathbf{A}_j-(\mathbf{B}\mathbf{PD})_j\|_2 \leq C_1 \varepsilon \ \ \text{for all $j \in [m]$}
\end{align}
%
for some permutation matrix $\mathbf{P}$ and invertible diagonal matrix $\mathbf{D}$.
%\footnote{We note that the condition $\varepsilon < L_2(\mathbf{A})$ above is necessary. When \mbox{$\mathbf{A}$ = $I$} and $\mathbf{x}_i = \mathbf{e}_i$, it turns out that $C_1 = 1$ and there is a $\mathbf{B}$ and $1$-sparse $\mathbf{\hat x}_i$ with $\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \varepsilon$ violating \eqref{Cstable}.}
Furthermore if $\varepsilon < L_{2k}(\mathbf{A}) / C_1$ then $\mathbf{B}$ also satisfies \eqref{SparkCondition} with $L_{2k}(\mathbf{B}\mathbf{PD}) \geq L_{2k}(\mathbf{A}) - C_1 \varepsilon$ and we in fact have:
\begin{align}\label{b-PDa}
\|\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\hat x}_i\|_1 &\leq  \left( \frac{ 1+C_1 \|\mathbf{x}_i\|_1 }{ L_{2k}(\mathbf{A}) -  C_1\varepsilon } \right) \varepsilon,
 \ \ \text{for all $i \in [N]$}.
\end{align}
\end{theorem}

For clarity of exposition, we delay explicitly defining the (effective) constant $C_1 = C_1(\mathbf{A}, \{\mathbf{x}_i\}_{i=1}^N, E)$ until Section \ref{DUT} (\eqref{Cdef}), since it requires additional definitions not critical to the statement of our results.

@@

Next, we address the case when only an upper bound $m'$ on the latent dimension $m$ is known (assuming that $\mathbf{B}$ satisfies \eqref{SparkCondition}).

\begin{theorem}\label{DeterministicUniquenessTheorem2}
Fix integers $n$ and $k < m \leq m'$ and matrices $\mathbf{A}~\in~\mathbb{R}^{n \times m}$ and $\mathbf{B} \in \mathbb{R}^{n \times m'}$ both satisfying \eqref{SparkCondition}. Suppose \mbox{$\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$} include at least \mbox{$(k-1){m' \choose k}+1$} $k$-sparse vectors in general linear position supported on each set in some $k$-uniform $E \subseteq 2^{[m]}$ satisfying the singleton intersection property. Then there exists a constant $C_3 > 0$ for which the following holds.

If for some $k$-sparse $\mathbf{\hat x}_1, \ldots, \mathbf{\hat x}_N \in \mathbb{R}^{m'}$ and $\varepsilon < L_2(\mathbf{A}) / C_3$ we have \mbox{$\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \varepsilon$} for all $i$, then:
\begin{align}\label{Cstablem'}
\|\mathbf{A}_j-(\mathbf{B}_J\mathbf{PD})_j\|_2 \leq C_3\varepsilon \ \ \text{for all $j \in [m]$}
\end{align}
%
for some $J \in {[m'] \choose m}$, permutation matrix $\mathbf{P}$, and invertible diagonal matrix $\mathbf{D}$.
\end{theorem}

In other words, the columns of $B$ contain (up to noise, after appropriate scaling) the columns of the original dictionary $\mathbf{A}$. Similarly, one can show by the same arguments at the beginning of Sec.~\ref{DUT} that the $\mathbf{\hat x}_i$ contain the original codes $\mathbf{x}_i$. The constant $C_3$ here is expression (\ref{Cdef2}) from the proof of Thm.~\ref{DeterministicUniquenessTheorem2}. Note that, in contrast to Thm.~\ref{DeterministicUniquenessTheorem}, this constant is dependent on $\mathbf{B}$; hence, \eqref{Cstablem'} holds for \emph{all} matrices $\mathbf{B} \in \mathbb{R}^{n \times m'}$ satisfying the spark condition only in the case $\varepsilon = 0$.

>>>>

Motivated by these discoveries and earlier work in the theory of neural communication \cite{Coulter10, Isely10}, we address when dictionaries and sparse representations are uniquely determined by data.  Answers to this question also have other real-world implications.  For example, a sparse coding analysis of local painting style can be used for forgery detection \cite{hughes2010, Olshausen10}, but only if sparse components of the artwork are independent of implementation idiosyncrasies. Fortunately, algorithms with proven recovery of generating dictionaries under certain 
conditions have recently been proposed (see \cite[Sec.~I-E]{Sun16} for a summary of the state-of-the-art).  However, there are few universal theorems explaining the general phenomenon of uniqueness.

@@

Note that any particular solution $(\mathbf{B}, \mathbf{\hat x}_1 \ldots, \mathbf{\hat x}_N)$ to this problem gives rise to an orbit of other solutions $(\mathbf{BPD}, \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\hat x}_1, \ldots, \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\hat x}_N)$, where $\mathbf{P}$ is any permutation matrix and $\mathbf{D}$ any invertible diagonal.  
We therefore consider whether solutions to Problem~\ref{InverseProblem} are unique only up to this ambiguity.  More specifically, 
we seek conditions for when sparsely coded datasets have a stable unique representation.

@@

%  We introduce the following terminology to handle $\eta > 0$. 
\begin{definition}\label{maindef}
Fix $Y = \{ \mathbf{y}_1, \ldots, \mathbf{y}_N\} \subset \mathbb{R}^n$. We say $Y$ has a \textbf{$k$-sparse linear representation in $m$ dimensions} if there exists a matrix $\mathbf{A}$ and $k$-sparse $\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$ such that $\mathbf{y}_i = \mathbf{A}\mathbf{x}_i$ for all $i$. %Equivalently, $\mathbf{Y} = \mathbf{A}\mathbf{X}$ for $\mathbf{Y}$ and $\mathbf{X}$ having the $\mathbf{y}_i$ and $\mathbf{x}_i$ as their columns, respectively. 
This representation is \textbf{stable} if for every $\delta_1, \delta_2 \geq 0$, there exists $\varepsilon = \varepsilon(\delta_1, \delta_2) \geq 0$ (with $\varepsilon > 0$ when  $\delta_1, \delta_2 > 0$) such that if a matrix $\mathbf{B}$ and $k$-sparse $\mathbf{\hat x}_1, \ldots, \mathbf{\hat x}_N \in \mathbb{R}^m$ satisfy:
\begin{align*}
\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \varepsilon, \ \ \text{for all $i$},
\end{align*}
%
then there is some permutation matrix $\mathbf{P}$ and invertible diagonal matrix $\mathbf{D}$ such that for all $i, j$:
\begin{align}\label{def1}
\|\mathbf{A}_j - (\mathbf{BPD})_j\|_2 \leq \delta_1 \ \ \text{and} \ \ \|\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\hat x}_i\|_1 \leq \delta_2.
\end{align}
\end{definition}
% Kato mentions a distance between n-tuples that minimizes the maximum difference over all possible orderings of one of the n-tuples.
% Make just 'sparse representation in m-dimensions? (i.e. with k unspecified?)

To see how Definition \ref{maindef} directly relates to Problem \ref{InverseProblem}, suppose that $Y$
% \mbox{$Y = \{\mathbf{A} \mathbf{x}_1, \ldots, \mathbf{A}\mathbf{x}_N\}$} 
has a stable $k$-sparse linear representation in $m$ dimensions and fix $\delta_1, \delta_2$ to be the desired accuracy in \eqref{def1}. Consider now any dataset $Z$ generated as in \eqref{LinearModel} with $\eta \leq \frac{1}{2} \varepsilon(\delta_1, \delta_2)$. Then, any dictionary $\mathbf{B}$ and $k$-sparse $\mathbf{\hat x}_1, \ldots, \mathbf{\hat x}_N$ solving Problem \ref{InverseProblem} are necessarily close to the original matrix $\mathbf{A}$ and codes $\mathbf{x}_i$ (i.e., satisfy \eqref{def1}).

@@

\begin{theorem}\label{DeterministicUniquenessTheorem}
% Fix positive integers $k, \ell, m, m', n, p$ satisfying $k < m$ and $p \leq m \leq m' \leq (\ell m - p) / (\ell - 1)$. 
Fix integers $n, \ell$ and $k < m \leq \hat m$. Suppose $\mathbf{A}~\in~\mathbb{R}^{n \times m}$ satisfies \eqref{SparkCondition} and that $k$-sparse \mbox{$\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$} contain at least \mbox{$(k-1){\hat m \choose k}+1$} vectors supported in general linear position on each set in some $\ell$-regular $E \subseteq {[m] \choose k}$ satisfying the SIP. Then there exists a constant $C_1 > 0$ for which the following holds for all $\varepsilon < L_2(\mathbf{A}) / C_1$.

Every matrix $\mathbf{B} \in \mathbb{R}^{n \times \hat m}$ for which for some $k$-sparse $\mathbf{\hat x}_1, \ldots, \mathbf{\hat x}_N \in \mathbb{R}^{\hat m}$ we have \mbox{$\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \varepsilon$} for all $i$ necessarily satisfies:
\begin{align}\label{Cstable}
\|\mathbf{A}_{\phi(j)} - (\mathbf{B}_{J}\mathbf{PD})_j\|_2 \leq C_1 \varepsilon \ \ \text{for $j = 1, \ldots, |J|$}
\end{align}
%
for some $J \subseteq [\hat m]$ of size \mbox{$\lfloor \hat m - \ell(\hat m - m) \rfloor$}, injective map $\phi$, permutation matrix $\mathbf{P}$ and invertible diagonal matrix $\mathbf{D}$.
%\footnote{We note that the condition $\varepsilon < L_2(\mathbf{A})$ above is necessary. When \mbox{$\mathbf{A}$ = $I$} and $\mathbf{x}_i = \mathbf{e}_i$, it turns out that $C_1 = 1$ and there is a $\mathbf{B}$ and $1$-sparse $\mathbf{\hat x}_i$ with $\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \varepsilon$ violating \eqref{Cstable}.}

Furthermore if $\varepsilon < L_{2k}(\mathbf{A}) / C_1$ then $\mathbf{B}_J$ also satisfies \eqref{SparkCondition} with $L_{2k}(\mathbf{B}_J\mathbf{PD}) \geq L_{2k}(\mathbf{A}) - C_1 \varepsilon$ and we in fact have [ ** TODO ** ]:
\begin{align}\label{b-PDa}
\|\mathbf{x}_i - \mathbf{D}^{-1}\mathbf{P}^{\top}\mathbf{\hat x}_i\|_1 &\leq  \left( \frac{ 1+C_1 \|\mathbf{x}_i\|_1 }{ L_{2k}(\mathbf{A}) -  C_1\varepsilon } \right) \varepsilon,
 \ \ \text{for all $i \in [N]$}.
\end{align}
\end{theorem}

In other words, the smaller the difference $\hat m - m$, the more columns and coefficients of the original $n \times m$ dictionary $\mathbf{A}$ and $m$-dimensional codes $\mathbf{x}_i$ contained (up to noise) in the appropriately scaled $n \times \hat m$ dictionary $\mathbf{B}$ and $\hat m$-dimensional codes $\mathbf{\hat x}_i$. In particular, if $\hat m = m$ then all columns of $\mathbf{A}$ and coefficients of each $\mathbf{x}_i$ recoverable, as per Def.~\ref{maindef}. For clarity of exposition, we delay explicitly defining the (effective) constant $C_1 = C_1(\mathbf{A}, \{\mathbf{x}_i\}_{i=1}^N, E)$ until Section \ref{DUT} (\eqref{Cdef}), since it requires additional definitions not critical to the statement of our results.

@@

%Next, we address the case when only an upper bound $m'$ on the latent dimension $m$ is known (assuming that $\mathbf{B}$ satisfies \eqref{SparkCondition}).

%\begin{theorem}\label{DeterministicUniquenessTheorem2}
%Fix integers $n$ and $k < m \leq m'$ and matrices $\mathbf{A}~\in~\mathbb{R}^{n \times m}$ and $\mathbf{B} \in \mathbb{R}^{n \times m'}$ both satisfying \eqref{SparkCondition}. Suppose \mbox{$\mathbf{x}_1, \ldots, \mathbf{x}_N \in \mathbb{R}^m$} include at least \mbox{$(k-1){m' \choose k}+1$} $k$-sparse vectors in general linear position supported on each set in some $k$-uniform $E \subseteq 2^{[m]}$ satisfying the singleton intersection property. Then there exists a constant $C_3 > 0$ for which the following holds.

%If for some $k$-sparse $\mathbf{\hat x}_1, \ldots, \mathbf{\hat x}_N \in \mathbb{R}^{m'}$ and $\varepsilon < L_2(\mathbf{A}) / C_3$ we have \mbox{$\|\mathbf{A}\mathbf{x}_i - \mathbf{B}\mathbf{\hat x}_i\|_2 \leq \varepsilon$} for all $i$, then:
%\begin{align}\label{Cstablem'}
%\|\mathbf{A}_j-(\mathbf{B}_J\mathbf{PD})_j\|_2 \leq C_3\varepsilon \ \ \text{for all $j \in [m]$}
%\end{align}
%
%for some $J \in {[m'] \choose m}$, permutation matrix $\mathbf{P}$, and invertible diagonal matrix $\mathbf{D}$.
%\end{theorem}

%In other words, the columns of $B$ contain (up to noise, after appropriate scaling) the columns of the original dictionary $\mathbf{A}$. Similarly, one can show by the same arguments at the beginning of Sec.~\ref{DUT} that the $\mathbf{\hat x}_i$ contain the original codes $\mathbf{x}_i$. The constant $C_3$ here is expression (\ref{Cdef2}) from the proof of Thm.~\ref{DeterministicUniquenessTheorem2}. Note that, in contrast to Thm.~\ref{DeterministicUniquenessTheorem}, this constant is dependent on $\mathbf{B}$; hence, \eqref{Cstablem'} holds for \emph{all} matrices $\mathbf{B} \in \mathbb{R}^{n \times m'}$ satisfying the spark condition only in the case $\varepsilon = 0$.
